### 2022-08-12 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@e6d22e1ebae71c288345b90ff61ad0c0da5a02be 🚀

We measured various properties of these configurations of maximum
entropy to evaluate how well they reproduce the structure of sampled
food webs. Specifically, we evaluated their nestedness \(\rho\), their
maximum trophic level \(maxtl\), their network diameter \(diam\), their
average maximum similarity between species pairs \(MxSim\) (Williams and
Martinez 2000), their proportion of cannibal species \(Cannib\), their
proportion of omnivorous species \(Omniv\), their SVD entropy, and their
motifs profile. Nestedness indicates how much the diet of specialist
species is a subset of the one of generalists (Delmas et al. 2019) and
was measured using the spectral radius of the adjacency matrix
(Staniczenko, Kopp, and Allesina 2013). In turn, the network diameter
represents the longest of the shortest paths between all species pair
(Albert and Barabasi 2002). Further, cannibal species are species that
can eat members of their own species (i.e.~species having self loops),
whereas omnivorous species can prey on different trophic levels
(Williams and Martinez 2000). Finally, a motifs profile represents the
proportion of three-species motifs (subnetworks), which can be
considered as simple building blocks of ecological networks (Milo et al.
2002; Stouffer et al. 2007). All of these properties are relatively easy
to measure and were chosen based on their ecological importance and
prevalent use in network ecology (McCann 2011; Delmas et al. 2019).

We compared the performance of both heuristic MaxEnt models in
predicting these measures to the one of the null
(tbl.~\ref{tbl:measures_all}) and neutral models
(tbl.~\ref{tbl:measures_abund}). Overall, we found that the models based
on the joint degree sequence (i.e.~the type II null and heuristic MaxEnt
models) reproduced the structure of empirical food webs much better than
the ones based on connectance (i.e.~the type I null and heuristic MaxEnt
models). This suggests that the predictive power of connectance might be
more limited than what was previously suggested (Timothée Poisot and
Gravel 2014). On the other hand, the neutral model of relative
abundances was surprisingly good at predicting the maximum trophic level
and the network diameter. However, with the exception of the network
diameter, the type II heuristic MaxEnt model was better at predicting
network structure than the neutral model for all measures considered.
This might be because, although neutral processes are important, they
act in concert with niche processes in determining species interactions
Canard et al. (2014). The joint degree sequence encodes information on
both neutral and niche processes because the number of prey and
predators a species has is determined by its relative abundance and
biological traits. Therefore, these results show that having information
on the number of prey and predators for each species substantially
improves the prediction of food-web structure, both compared to models
solely based on connectance and the ones solely based on species
relative abundances.

Next, the predictions of the type II heuristic MaxEnt model can be
compared to its null model counterpart. On average, the type II
heuristic MaxEnt model was better at predicting nestedness
(\(0.62 \pm 0.08\)) than its corresponding null model
(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for all
networks in our complete dataset (tbl.~\ref{tbl:measures_all}). This
might in part be due to the fact that nestedness was calculated using
the spectral radius of the adjacency matrix, which directly leverages
information on the network itself just like the heuristic MaxEnt models.
The proportion of self-loops (cannibal species) was also better
predicted by the type II heuristic MaxEnt model in comparison to the
type II null model. However, the type II null model was better at
predicting network diameter and average maximum similarity between
species pairs, and predictions of the maximum trophic level and the
proportion of omnivorous species were similar between both types of
models. This suggests that, although the joint degree sequence is an
important driver of food-web structure, other ecological constraints
might need to be added to our MaxEnt model to account for some emerging
food-web properties, especially the ones regarding food-chain lengths.

@@

Finally, in fig.~\ref{fig:measures} we also show how well empirical
measures are predicted using the type II MaxEnt network model. In
accordance with our previous results, we found that nestedness was very
well predicted by our model. However, the type II MaxEnt network model
overestimated the maximum trophic level and network diameter, especially
for networks with high empirical values. In fig.~S3, we show that the
pairwise relationships between these four measures and species richness
in empirical food webs are similar (in magnitude and sign) to the ones
in type II MaxEnt networks.

\begin{figure}
\hypertarget{fig:measures}{%
\centering
\includegraphics{figures/measures_emp_maxent.png}
\caption{Relationship between the structure of empirical and maximum
entropy food webs. Empirical networks include most food webs archived on
Mangal, as well as the New Zealand and Tuesday Lake datasets (our
complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the joint degree sequence (type II MaxEnt network
model). (a) Nestedness (estimated using the spectral radius of the
adjacency matrix), (b) the maximum trophic level, (c) the network
diameter, and (d) the SVD entropy were measured on these empirical and
maximum entropy food webs. The identity line is plotted in each
panel.}\label{fig:measures}
}
\end{figure}

Moreover, we found that empirical networks are close to their maximum
entropy considering a fixed joint degree sequence (fig.~S4). Empirical
networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
between the SVD entropy of empirical food webs and their maximum entropy
configuration is plotted in the last panel of fig.~\ref{fig:measures}.
We also found no relationship between the difference in SVD entropy
between empirical and type II MaxEnt networks and species richness, the
number of interactions, or connectance (fig.~S5). Similarly, we found no
correlation between the difference in SVD entropy and the Jaccard
distance of empirical and type II MaxEnt networks.

The picture slightly changes when we consider another important property
of ecological networks, i.e.~their motifs profile. We measured the
proportion of three-species motifs, which can be considered as simple
building blocks of ecological networks (Milo et al. 2002; Stouffer et
al. 2007). Motifs are the backbone of complex ecological networks from
which network structure is built upon and play a crucial role in
community dynamics and assembly (Stouffer and Bascompte 2011). In
fig.~\ref{fig:motifs}, we show that the motifs profile of networks
generated using the type II MaxEnt network model was very close to the
one of empirical data, and that they made better predictions than the
type II null model. This is also shown in fig.~\ref{fig:motifs_rel},
where we see that relationships between motifs pairs of empirical food
webs are very similar to the ones of the type II MaxEnt networks.
Overall, these results suggest that the type II MaxEnt network model can
reproduce many aspects of network structure. This highlights the
importance of the joint degree sequence in shaping ecological networks.

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include most food webs archived on Mangal, as
well as the New Zealand and Tuesday Lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulated
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include most food webs
archived on Mangal, as well as the New Zealand and Tuesday Lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the connectance (type I MaxEnt network model) or the
joint degree sequence (type II MaxEnt network model). The predictions of
the types I and II null models are also plotted. Regression lines are
plotted in each panel. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs_rel}
}
\end{figure}

>>>>

We measured various properties of these configurations of maximum
entropy to evaluate how well they reproduce the structure of sampled
food webs. Specifically, we evaluated their nestedness \(\rho\), their
maximum trophic level \(maxtl\), their network diameter \(diam\), their
average maximum similarity between species pairs \(MxSim\) (Williams and
Martinez 2000), their proportion of cannibal species \(Cannib\), their
proportion of omnivorous species \(Omniv\), their SVD entropy, and their
motifs profile. Nestedness indicates how much the diet of specialist
species is a subset of the one of generalists (Delmas et al. 2019) and
was measured using the spectral radius of the adjacency matrix
(Staniczenko, Kopp, and Allesina 2013). In turn, the network diameter
represents the longest of the shortest paths between all species pair
(Albert and Barabasi 2002). Further, cannibal species are species that
can eat individuals of their own species (i.e.~species having self
loops), whereas omnivorous species can prey on different trophic levels
(Williams and Martinez 2000). Finally, a motifs profile represents the
proportion of three-species motifs (subnetworks), which can be
considered as simple building blocks of ecological networks (Milo et al.
2002; Stouffer et al. 2007). All of these properties are relatively easy
to measure and were chosen based on their ecological importance and
prevalent use in network ecology (McCann 2011; Delmas et al. 2019).

We compared the performance of both heuristic MaxEnt models in
predicting these measures to the one of the null
(tbl.~\ref{tbl:measures_all}) and neutral models
(tbl.~\ref{tbl:measures_abund}). Overall, we found that the models based
on the joint degree sequence (i.e.~the type II null and heuristic MaxEnt
models) reproduced the structure of empirical food webs much better than
the ones based on connectance (i.e.~the type I null and heuristic MaxEnt
models). This suggests that the predictive power of connectance might be
more limited than what was previously suggested (Timothée Poisot and
Gravel 2014). On the other hand, the neutral model of relative
abundances was surprisingly good at predicting the maximum trophic level
and the network diameter. However, with the exception of the network
diameter, the type II heuristic MaxEnt model was better at predicting
network structure than the neutral model for most measures considered.
This might be because, although neutral processes are important, they
act in concert with niche processes in determining species interactions
Canard et al. (2014). The joint degree sequence encodes information on
both neutral and niche processes because the number of prey and
predators a species has is determined by its relative abundance and
biological traits. These results thus show that having information on
the number of prey and predators for each species substantially improves
the prediction of food-web structure, both compared to models solely
based on connectance and to the ones solely based on species relative
abundances.

Next, the predictions of the type II heuristic MaxEnt model can be
compared to its null model counterpart. On average, the type II
heuristic MaxEnt model was better at predicting nestedness
(\(0.62 \pm 0.08\)) than its corresponding null model
(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for networks
in our complete dataset (tbl.~\ref{tbl:measures_all}). This might in
part be due to the fact that nestedness was calculated using the
spectral radius of the adjacency matrix, which directly leverages
information on the network itself just like the heuristic MaxEnt model.
The proportion of self-loops (cannibal species) was also better
predicted by the type II heuristic MaxEnt model in comparison to the
type II null model. However, the type II null model was better at
predicting network diameter and average maximum similarity between
species pairs, and predictions of the maximum trophic level and the
proportion of omnivorous species were similar between both types of
models. We believe that this is because increasing the complexity of a
food web might increase its average and maximum food-chain lengths. In
comparison, the null model was more stochastic and does not necessarily
produce more complex food webs with longer food-chain lengths.

@@

Despite this increase in maximum trophic level and network diameter in
MaxEnt food webs, we found that empirical food webs are close to their
maximum entropy given a fixed joint degree sequence (fig.~S3). Empirical
food webs in the complete dataset had an SVD entropy of
\(0.89 \pm 0.04\), compared to an SVD entropy of \(0.94 \pm 0.03\) for
networks generated using the type II heuristic MaxEnt model. The
relationship between the SVD entropy of empirical food webs and their
maximum entropy is plotted in the last panel of fig.~\ref{fig:measures}.
As expected, the SVD entropy of maximum entropy food webs was higher
than that of empirical food webs for almost all networks, confirming
that our method indeed generated more complex networks. Moreover, we
found no to a weak relationship between the increase in SVD entropy and
the number of species, the number of interactions, and connectance
(fig.~S4). This suggests that the slight increase in entropy between
empirical food webs and their maximum entropy configuration was
sufficient to modify some of their properties, regardless of their
number of species and their number of interactions.

A direct comparison of the structure of maximum entropy food webs,
constrained by the joint degree sequence, with empirical data also
supports the results depicted in tbl.~\ref{tbl:measures_all}. Indeed, in
fig.~\ref{fig:measures} we show how well empirical measures are
predicted by the type II heuristic MaxEnt model. In accordance with our
previous results, we found that nestedness was very well predicted by
our model. However, the model overestimated the maximum trophic level
and network diameter, especially when the sampled food web had
intermediate values of these measures. In fig.~S5, we show that the
pairwise relationships between the four measures in
fig.~\ref{fig:measures} and species richness in empirical food webs are
similar (in magnitude and sign) to the ones found in food webs generated
using the type II heuristic MaxEnt model. This indicates that the number
of species in the network does not seem to impact the ability of the
model to reproduce food-web structure.

\begin{figure}
\hypertarget{fig:measures}{%
\centering
\includegraphics{figures/measures_emp_maxent.png}
\caption{Relationship between the structure of empirical and maximum
entropy food webs. Empirical networks include most food webs archived on
Mangal, as well as the New Zealand and Tuesday Lake food webs (our
complete dataset). Maximum entropy networks were obtained using the type
II heuristic MaxEnt model based on the joint degree sequence. (a)
Nestedness (estimated using the spectral radius of the adjacency
matrix), (b) the maximum trophic level, (c) the network diameter, and
(d) the SVD entropy were measured on these empirical and maximum entropy
food webs. The identity line is plotted in each
panel.}\label{fig:measures}
}
\end{figure}

Notwithstanding its difficulties in reproducing adequate measures of
food-chain lengths, the type II heuristic MaxEnt model can predict
surprisingly well the motifs profile. Motifs are the backbone of complex
ecological networks from which network structure is built upon and play
a crucial role in community dynamics and assembly (Stouffer and
Bascompte 2011). For this reason, the motifs profile can act as an
effective ecological constraint shaping species interactions networks,
and thus constitute a substantial source of ecological information. In
fig.~\ref{fig:motifs}, we show that the motifs profile of networks
generated using the type II heuristic MaxEnt model accurately reproduced
the one of empirical data. This model made significantly better
predictions than the ones based on connectance and the type II null
model based on the joint degree sequence. This is also shown in
fig.~\ref{fig:motifs_rel}, where we see that the relationships between
motifs proportions in empirical food webs are very similar to the ones
in networks generated using the type II heuristic MaxEnt model. This is
in contrast with the type I null and MaxEnt models based on connectance,
which produced opposite relationships than what was observed
empirically. Our findings thus suggest that increasing food-web
complexity within a maximum entropy framework based on the joint degree
sequence does not alter the proportion of three-species motifs, but
might alter their position relative to one another.

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical and predicted food webs. Empirical
networks include most food webs archived on Mangal, as well as the New
Zealand and Tuesday Lake datasets (our complete dataset). Null 1: Type I
null model based on connectance. MaxEnt 1: Type I heuristic MaxEnt model
based on connectance. Null 2: Type II null model based on the joint
degree sequence. MaxEnt 2: Type II heuristic MaxEnt model based on the
joint degree sequence. Boxplots display the median proportion of each
motif (middle horizontal lines), as well as the first (bottom horizontal
lines) and third (top horizontal lines) quartiles. Vertical lines
encompass all data points that fall within \(1.5\) times the
interquartile range from both quartiles, and dots are data points that
fall outside this range. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Pairwise relationships between motifs proportions of empirical
and predicted food webs. Empirical networks include most food webs
archived on Mangal, as well as the New Zealand and Tuesday Lake datasets
(our complete dataset). Null 1: Type I null model based on connectance.
MaxEnt 1: Type I heuristic MaxEnt model based on connectance. Null 2:
Type II null model based on the joint degree sequence. MaxEnt 2: Type II
heuristic MaxEnt model based on the joint degree sequence. Regression
lines are plotted in each panel. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs_rel}
}
\end{figure}

One of the challenges in implementing and validating a maximum entropy
model is to discover where its predictions break down. The results
depicted in tbl.~\ref{tbl:measures_all} and fig.~\ref{fig:measures} show
that our type II heuristic MaxEnt model can capture many high-level
properties of food webs, but does a poor job of capturing others. This
suggests that, although the joint degree sequence is an important driver
of food-web structure, other ecological constraints might be needed to
account for some emerging food-web properties, especially the ones
regarding food-chain lengths. Nevertheless, fig.~\ref{fig:motifs} and
fig.~\ref{fig:motifs_rel} show that the model can reproduce surprisingly
well the motifs profile, one of the most ecologically informative
properties of food webs. This suggests that the emerging structure of
food webs is mainly driven by their joint degree sequence, although
higher-level properties might be needed to ensure that food-chain
lengths fall within realistic values.

### 2022-08-12 suppmat.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@fa572cda4b73fa82f7f24e762cbdf281e286e661 🚀
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@e6d22e1ebae71c288345b90ff61ad0c0da5a02be 🚀

\begin{figure}
\hypertarget{fig:entropy_dist}{%
\centering
\includegraphics{figures/entropy_distribution.png}
\caption{(a) Distribution of the SVD entropy of MaxEnt food webs (type
II MaxEnt network model) and of empirical food webs. Empirical networks
include most food webs archived on Mangal, as well as the New Zealand
and Tuesday lake datasets (our complete dataset). Maximum entropy
networks were derived using a simulating annealing algorithm to find the
network of maximum SVD entropy while maintaining the joint degree
sequence. (b) Distribution of z-scores of the SVD entropy of all
empirical food webs. Z-scores were computed using the mean and standard
deviation of the distribution of SVD entropy of MaxEnt food webs (type
II MaxEnt network model). The dash line corresponds to the median
z-score.}\label{fig:entropy_dist}
}
\end{figure}

\begin{figure}
\hypertarget{fig:entropy_size}{%
\centering
\includegraphics{figures/difference_entropy.png}
\caption{Difference in SVD entropy between MaxEnt (type II MaxEnt
network model) and empirical food webs as a function of (a) species
richness, (b) the number of interactions, and (c) connectance. Empirical
networks include most food webs archived on Mangal, as well as the New
Zealand and Tuesday lake datasets (our complete dataset). Maximum
entropy networks were derived using a simulating annealing algorithm to
find the network of maximum SVD entropy while maintaining the joint
degree sequence. Regression lines are plotted in each
panel.}\label{fig:entropy_size}
}
\end{figure}

\begin{figure}
\hypertarget{fig:measures_richness}{%
\centering
\includegraphics{figures/measures_richness.png}
\caption{Structure of empirical and maximum entropy food webs (type II
MaxEnt network model) as a function of species richness. Empirical
networks include most food webs archived on Mangal, as well as the New
Zealand and Tuesday lake datasets (our complete dataset). Maximum
entropy networks were derived using a simulating annealing algorithm to
find the network of maximum SVD entropy while maintaining the joint
degree sequence. (a) Nestedness (estimated with the spectral radius of
the adjacency matrix), (b) the maximum trophic level, (c) the network
diameter (i.e., the longest shortest path between all species pairs),
and (d) the SVD entropy were measured on these empirical and predicted
food webs and plotted against species richness. Regression lines are
plotted in each panel.}\label{fig:measures_richness}
}
\end{figure}

>>>>

\begin{figure}
\hypertarget{fig:entropy_dist}{%
\centering
\includegraphics{figures/entropy_distribution.png}
\caption{(a) Distribution of the SVD entropy of empirical and maximum
entropy food webs. Empirical networks include most food webs archived on
Mangal, as well as the New Zealand and Tuesday Lake food webs (our
complete dataset). Maximum entropy networks were obtained using the type
II heuristic MaxEnt model based on the joint degree sequence. (b)
Distribution of z-scores of the SVD entropy of all empirical food webs.
Z-scores were computed using the mean and standard deviation of the
distribution of SVD entropy of MaxEnt food webs (type II heuristic
MaxEnt model). The dash line corresponds to the median
z-score.}\label{fig:entropy_dist}
}
\end{figure}

\begin{figure}
\hypertarget{fig:entropy_size}{%
\centering
\includegraphics{figures/difference_entropy.png}
\caption{Difference in SVD entropy between maximum entropy and empirical
food webs as a function of (a) species richness, (b) the number of
interactions, and (c) connectance. Empirical networks include most food
webs archived on Mangal, as well as the New Zealand and Tuesday Lake
food webs (our complete dataset). Maximum entropy networks were obtained
using the type II heuristic MaxEnt model based on the joint degree
sequence. Regression lines are plotted in each
panel.}\label{fig:entropy_size}
}
\end{figure}

\begin{figure}
\hypertarget{fig:measures_richness}{%
\centering
\includegraphics{figures/measures_richness.png}
\caption{Structure of empirical and maximum entropy food webs as a
function of species richness. Empirical networks include most food webs
archived on Mangal, as well as the New Zealand and Tuesday Lake food
webs (our complete dataset). Maximum entropy networks were obtained
using the type II heuristic MaxEnt model based on the joint degree
sequence. (a) Nestedness (estimated using the spectral radius of the
adjacency matrix), (b) the maximum trophic level, (c) the network
diameter, and (d) the SVD entropy were measured on these empirical and
maximum entropy food webs and plotted against species richness.
Regression lines are plotted in each
panel.}\label{fig:measures_richness}
}
\end{figure}
